---
title: "dsp_bac_host_cor"
output: html_document
---

This script calcualtes the correlation between genes from Ecad ROIs and bacteria signals from fecal ROIs. Note the correlations are calculated between paired (adjacent) Ecad-fecal ROIs.


```{r}
############### read in dsp data host

dsp_counts = read.csv('../../dsp_related/final_data/microdsp_q3Counts.csv') # host DSP Q3 norm counts

################ bacteria version
## read in dsp data
bac_counts = read.csv('../../dsp_related/final_data/microdsp_q3V2Counts.csv') # same but bacteria counts
###
meta = read.table("../../dsp_related/final_data/meta_analys_use.txt", sep = '\t', header = T, na.strings=c("","NA"))
meta_host = meta[meta$slide.name != '0117-BCR' & # remove ROIs from BMA validation
                                  #meta$segment != 'Full ROI' &
                                  meta$slide.name != 'No Template Control',]# sample names for wta analysis
meta_host$treatment = substr(meta_host$tag2, 0, 1)

library(stringr)
full_meta = read.csv('../../../meta_link/meta_mibi_dsp.csv', na.strings=c("","NA"))
colnames(full_meta)[2] = "run"
colnames(full_meta)[3] = "fov"

full_meta_bacteria = full_meta[!is.na(full_meta$DSP_Bac_name),]
indexget = full_meta_bacteria$DSP_Bac_name %in% dsp_counts$X & full_meta_bacteria$DSP_Ecad_name %in% dsp_counts$X
full_meta_bacteriaf = full_meta_bacteria[indexget,]
# quick check if regions in data
full_meta_bacteriaf$treatment = substr(full_meta_bacteriaf$run, 24, 24)


#### change row names for dsp host and bacteria coutns

rownames(dsp_counts) = dsp_counts$X
dsp_counts$X = NULL
rownames(bac_counts) = bac_counts$X
bac_counts$X = NULL
```


######## First thing to test:

dss: large: bac vs Ecad
ctrl: large: bac vs Ecad

```{r}
# dss: large: bac vs Ecad
full_meta_bacteriaf_dss = subset(full_meta_bacteriaf, full_meta_bacteriaf$treatment == 'T' &
                                   full_meta_bacteriaf$location == 'Large')
x_n = meta_host$Sample_ID[match(full_meta_bacteriaf_dss$DSP_Bac_name, meta_host$Sample_ID)] # bac
y_n = meta_host$Sample_ID[match(full_meta_bacteriaf_dss$DSP_Ecad_name, meta_host$Sample_ID)] # ecad

x = bac_counts[x_n,]
y = dsp_counts[y_n,c(1:8887)]

### corr
library(CCA)
xy = matcor(x,y)
corr_xy = xy$XYcor[c(1:4),c(5:8891)] # corr between bacteria 

```


##### first get host genes that are highly correlated with adjacnet bacteria signals (Top 50)

```{r}
### plotting all genes seems too ambitious, lets get head and tails
temp = list()
for (item in rownames(corr_xy)){
  temp = append(temp, colnames(corr_xy)[order(corr_xy[item,], decreasing = T)[c(1:50)]])
  #temp = append(temp, colnames(corr_xy)[order(corr_xy[item,], decreasing = F)[c(1:20)]])
}

plt_genes = unique(unlist(temp))
plt_genes

```



```{r}
library(viridis)
#svg(file="../plots/bac_host_corr_dsp_ht50_nolowCorr_colortest.svg",height = 4, width = 20)
#pheatmap(corr_xy[c('Firm', 'Bac', 'Pro'),plt_genes], color = inferno(50))
#dev.off()

##
pdf(file="../plots/bac_host_corr_dsp_ht50_nolowCorr_colortest2.pdf",height = 3, width = 20)
pheatmap(corr_xy[c('Firm', 'Bac', 'Pro'),plt_genes], color = viridis(50),
         treeheight_row = 0, treeheight_col = 0)
dev.off()

##
#tiff(file="../plots/bac_host_corr_dsp_ht50_nolowCorr_colortest.tiff",height = 4, width = 20)
#pheatmap(corr_xy[c('Firm', 'Bac', 'Pro'),plt_genes], color = inferno(50))
#dev.off()
```



##### then we can perform gene ontology analysis based on the highly correlated genes

```{r}
library(enrichR)
setEnrichrSite("Enrichr") #human or mouse db
dbs <- c("GO_Biological_Process_2015") # consistent with previous analysis
```

## we found top correlated genes are all metabolic processes:

```{r}
##
algg = colnames(corr_xy)[order(corr_xy['Firm',], decreasing = T)[c(1:50)]]
enriched <- enrichr(algg, dbs)
temp = enriched$GO_Biological_Process_2015
print(temp$Term[c(1:10)])

## important genes
#phosphatidylethanolamine biosynthetic process (GO:0006646) - SLC27A1;PHOSPHO1
#fatty acid metabolic process (GO:0006631) - SLC27A1;EHHADH;MGLL;ACSF2;SLC27A4
```

```{r}
##
algg = colnames(corr_xy)[order(corr_xy['Bac',], decreasing = T)[c(1:50)]]
enriched <- enrichr(algg, dbs)
temp = enriched$GO_Biological_Process_2015
print(temp$Term[c(1:10)])

## important genes
#	membrane lipid metabolic process (GO:0006643) - GAL3ST1;A4GALT;SMPDL3A;ST6GALNAC4;SERINC5
#lipid biosynthetic process (GO:0008610) - FADS2;GAL3ST1;ACSS2;ELOVL7;A4GALT;SERINC5
```

```{r}
##
algg = colnames(corr_xy)[order(corr_xy['Pro',], decreasing = T)[c(1:50)]]
enriched <- enrichr(algg, dbs)
temp = enriched$GO_Biological_Process_2015
print(temp$Term[c(1:10)])

## genes that are important:
#phosphatidic acid biosynthetic process (GO:0006654) - PLA2G2A;PLA2G10;AGPAT1
#cholesterol metabolic process (GO:0008203) - CELA3B;EBP;GGPS1;CEL
#intestinal cholesterol absorption (GO:0030299) - CEL;PNLIP
```



