---
title: "mibi_cluster_final"
output: html_document
---

In this script we perform cell type annotation view seurat louvain clustering. Also contains cell type related analysis and plotting presented in figure 3.

```{r}
# two runs combine them
df1 = read.csv('../data/celldf_batch1.csv')
df2 = read.csv('../data/celldf_batch2.csv')
df12 = rbind(df1, df2) # 130k unfiltered cells
```


```{r}
# first check cell size
ggplot(df12, aes(x=size)) + geom_histogram(binwidth = 20)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))
```

```{r}
## too large cell removed
## based on this plot the small cells should be <= 50 pixels
df12 = df12[df12$size<=2000,]
print(dim(df12))
ggplot(df12, aes(x=size)) + geom_histogram(binwidth = 10)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 20))
```

```{r}
# remove small cell dirts
## based on this plot the small cells should be <= 50 pixels
df12 = df12[df12$size>=50,]
print(dim(df12))
ggplot(df12, aes(x=size)) + geom_histogram(binwidth = 10)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 20))
```

```{r}
# check based on dsdna expression density
df12$dsDNA_host_density = df12$dsDNA_host/ df12$size
ggplot(df12, aes(x=dsDNA_host_density)) + geom_histogram(binwidth = 0.01)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 20))
```

```{r}
# remove extreme bright cells
df12 = df12[df12$dsDNA_host_density<=0.7,]
print(dim(df12))
ggplot(df12, aes(x=dsDNA_host_density)) + geom_histogram(binwidth = 0.001)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 25))
```

```{r}
# remove extreme dim cells
df12 = df12[df12$dsDNA_host_density>=0.01,]
print(dim(df12))
ggplot(df12, aes(x=dsDNA_host_density)) + geom_histogram(binwidth = 0.001)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 25))
```


```{r}
# then normalization
col_use = c("B220", "CD3e", "CD4", "CD11b", "CD11c", "CD31", "CD45", "CD68", "DCAMLK1",
            "Ecad", "F480", "IgA", "Ki67", "Ly6g", "MUC2", "PNAD.1", "Reg3beta", "SMA", "Tubulin", "Vimentin")
df12_marker = df12[, col_use]
df12_marker = log1p(df12_marker) # log1p norm
```


Our clustering strategy:
sepearte immune and non-immune cells by CD45 expression
cluster seperately.


```{r}
# look at cd45 signal distribution
temp = df12_marker
temp$CD45[temp$CD45>2] = 2
quantile(temp$CD45,0.9) #1.15
ggplot(temp, aes(x=CD45)) + geom_histogram(binwidth = 0.01)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 25))
```

```{r}
# quick check cd45 indicator

run = '2023-03-20T14-32-22-DSST1-Run2' #'2023-03-05T18-22-42-DSST2-Run2' # '2023-03-08T18-48-43-DSST3-Run5'
fov = 'fov-1-scan-1' #'fov-2-scan-1'#'fov-1-scan-1'

plotsub = df12[(df12$run == run) & (df12$fov == fov),]
plotsub$CD45_ind = '0'
plotsub$CD45_ind[plotsub$CD45>=1.15] = '1'

ggplot(plotsub, aes(x = y, y = -x, color = CD45_ind)) + geom_point(size = 2) + theme_classic() + 
  scale_color_manual(values=as.vector(polychrome(30))) # this looks convincing
```


```{r}
# gate them into two parts, cluster them seperately
df12_marker_cd45 = df12_marker[df12_marker$CD45 >= 1.15,]
df12_marker_other = df12_marker[df12_marker$CD45 < 1.15,]


# cluster cd45 first
col_use = c("B220", "CD3e", "CD4", "CD11b", "CD11c", "CD68", "F480", "IgA", "Ly6g")
use = df12_marker_cd45[, col_use]
library(Seurat)
tm=as.data.frame(t(use))

# seurat clustering
micromibi_cd45 <- CreateSeuratObject(counts =tm)
micromibi_cd45 <- SetAssayData(object = micromibi_cd45, slot = "data", new.data = as.matrix(tm)) # no need since already log1p
micromibi_cd45 <- SetAssayData(object = micromibi_cd45, slot = "scale.data", new.data = as.matrix(tm)) # no scale needed for our custom case
micromibi_cd45 <- RunPCA(micromibi_cd45, features = rownames(micromibi_cd45))

micromibi_cd45 <- RunUMAP(micromibi_cd45, dims = 1:8)
micromibi_cd45 <- FindNeighbors(micromibi_cd45, features = rownames(micromibi_cd45), dims = 1:8)
micromibi_cd45 <- FindClusters(micromibi_cd45, resolution = 0.3)
DimPlot(micromibi_cd45, label = TRUE) #+ NoLegend()
```

```{r}
pltdf_clst = df12_marker_cd45#[,markers_use]
pltdf_clst[,col_use] = scale(pltdf_clst[,col_use])
pltdf_clst$Meta = Idents(micromibi_cd45)
# calculate heatmap from all clusters - by Meta
pltdf_clst %>%
  dplyr::group_by(Meta) %>%
  dplyr::summarise_if(is.numeric, median) %>%
  ungroup() ->
  heat_f
rownames(heat_f) <- heat_f$Meta
#table(fsom_clustering_data$Meta)
## manual cap
cc = colnames(heat_f)[2:21]
upbd = 1.5
lwbd = -1.5
for (c in cc){
  heat_f[heat_f[,c]>upbd,c] = upbd
  heat_f[heat_f[,c]<lwbd,c] = lwbd
}

rownames(heat_f) = as.character(as.integer(rownames(heat_f)) - 1)
heatmaply(heat_f[,col_use], labRow = rownames(heat_f), main="Meta",
          scale_fill_gradient_fun = scale_fill_gradient2(low="blue",high="red",midpoint=0,limits=c(lwbd,upbd))
          )
```

Immune clusters:

0 DC
1 Macrophage
2 Macrophage
3 Plasma
4 B
5 Monocyte
6 T
7 Other Immune
8 B
9 Neutrophil

Name the clusters

```{r}
seurat_labels_imm = Idents(micromibi_cd45)
# manual annotation
new.cluster.ids <- c("DC","Macrophage","Macrophage","Plasma","B","Monocyte","T",
                     "Other Immune","B","Neutrophil")

names(new.cluster.ids) <- levels(micromibi_cd45)
micromibi_cd45 <- RenameIdents(micromibi_cd45, new.cluster.ids)
table(Idents(micromibi_cd45))
```


```{r}
DimPlot(micromibi_cd45, label = TRUE) #+ NoLegend()
```


############# cluster out non-immune cells


```{r}
# cluster non-immune cells then

col_use = c("Ecad", "Ki67", "MUC2", "PNAD.1", "SMA", "Tubulin", "Vimentin", "CD31")

use = df12_marker_other[, col_use]
library(Seurat)
tm=as.data.frame(t(use))

# seurat clustering
micromibi_other <- CreateSeuratObject(counts =tm)
micromibi_other <- SetAssayData(object = micromibi_other, slot = "data", new.data = as.matrix(tm)) # no need since already log1p
micromibi_other <- SetAssayData(object = micromibi_other, slot = "scale.data", new.data = as.matrix(tm)) # no scale needed for our custom case
micromibi_other <- RunPCA(micromibi_other, features = rownames(micromibi_other))

micromibi_other <- RunUMAP(micromibi_other, dims = 1:7)
micromibi_other <- FindNeighbors(micromibi_other, features = rownames(micromibi_other), dims = 1:7)
micromibi_other <- FindClusters(micromibi_other, resolution = 0.3)
DimPlot(micromibi_other, label = TRUE) #+ NoLegend()
```


```{r}
pltdf_clst = df12_marker_other#[,markers_use]
pltdf_clst[,col_use] = scale(pltdf_clst[,col_use])
pltdf_clst$Meta = Idents(micromibi_other)
# calculate heatmap from all clusters - by Meta
pltdf_clst %>%
  dplyr::group_by(Meta) %>%
  dplyr::summarise_if(is.numeric, median) %>%
  ungroup() ->
  heat_f
rownames(heat_f) <- heat_f$Meta
#table(fsom_clustering_data$Meta)
## manual cap
cc = colnames(heat_f)[2:21]
upbd = 1.5
lwbd = -1.5
for (c in cc){
  heat_f[heat_f[,c]>upbd,c] = upbd
  heat_f[heat_f[,c]<lwbd,c] = lwbd
}
rownames(heat_f) = as.character(as.integer(rownames(heat_f)) - 1)
heatmaply(heat_f[,col_use], labRow = rownames(heat_f), main="Meta",
          scale_fill_gradient_fun = scale_fill_gradient2(low="blue",high="red",midpoint=0,limits=c(lwbd,upbd))
          )
```

Other clusters:

0 Epithelial
1 Goblet
2 Goblet
3 Smooth muscle (Ki67 high)
4 Epithelial
5 Smooth muscle
6 Epithelial (Muc2+)
7 Epithelial
8 Epithelial (Muc2+)
9 Smooth muscle
10 Epithelial

Name the clusters

```{r}
seurat_labels_other = Idents(micromibi_other)
# manual annotation
#new.cluster.ids <- c("Goblet","Epithelial","Smooth muscle","Epithelial","Epithelial","Smooth muscle (Ki67 high)","Epithelial",
#                     "Epithelial","Epithelial","Epithelial")

new.cluster.ids <- c("Epithelial","Goblet","Goblet","Smooth muscle (Ki67 high)","Epithelial","Smooth muscle","Epithelial (Muc2+)",
                     "Epithelial","Epithelial (Muc2+)","Smooth muscle", "Epithelial")

names(new.cluster.ids) <- levels(micromibi_other)
micromibi_other <- RenameIdents(micromibi_other, new.cluster.ids)
table(Idents(micromibi_other))
```

```{r}
DimPlot(micromibi_other, label = TRUE) #+ NoLegend()
```


Okay now we can put in the full list and visualize


```{r}
col_use = c("B220", "CD3e", "CD4", "CD11b", "CD11c", "CD31", "CD45", "CD68",
            "Ecad", "F480", "IgA", "Ki67", "Ly6g", "MUC2", "PNAD.1", "SMA", "Tubulin", "Vimentin")
df12_marker = df12[, col_use]
df12_marker = log1p(df12_marker) # log1p norm

library(Seurat)
tm=as.data.frame(t(df12_marker))
micromibi <- CreateSeuratObject(counts =tm)
micromibi <- SetAssayData(object = micromibi, slot = "data", new.data = as.matrix(tm)) # no need since already log1p
micromibi <- SetAssayData(object = micromibi, slot = "scale.data", new.data = as.matrix(tm)) # no scale needed for our custom case
micromibi <- RunPCA(micromibi, features = rownames(micromibi))
micromibi <- RunUMAP(micromibi, dims = 1:15)

micromibi@meta.data$annotationV1 = "None"
micromibi@meta.data$annotationV1[df12_marker$CD45 >= 1.15] = as.character(Idents(micromibi_cd45))
micromibi@meta.data$annotationV1[df12_marker$CD45 < 1.15] = as.character(Idents(micromibi_other))
micromibi@meta.data$annotationV1 <- factor(micromibi@meta.data$annotationV1,
                                           levels = c("Epithelial", "Epithelial (Muc2+)" ,"Goblet", "Smooth muscle", "Smooth muscle (Ki67 high)",
                                                      "B", 'Plasma',"T", "Monocyte", "Macrophage", 'DC', 'Other Immune',
                                                      'Neutrophil'))

DimPlot(micromibi, group.by = 'annotationV1', raster = F) + #
  scale_color_manual(values = c("#8DD3C7","#BEBADA", "#FB8072", "#A6CEE3",
                                "#1F78B4", "#B2DF8A", "#33A02C",
                                "#FB9A99", "#E31A1C", "#FDBF6F",
                                "#FF7F00", "#1B9E77", "#D95F02"))#scale_color_npg() #+ NoLegend() 
```


```{r}
#save out the final version
svdf = df12

# we only use the normalized values
col_use = c("B220", "CD3e", "CD4", "CD11b", "CD11c", "CD31", "CD45", "CD68", "DCAMLK1",
            "Ecad", "F480", "IgA", "Ki67", "Ly6g", "MUC2", "PNAD.1", "Reg3beta", "SMA", "Tubulin", "Vimentin")
df12_marker = df12[, col_use]
df12_marker = log1p(df12_marker) # log1p norm

svdf[,col_use] = df12_marker
svdf$annV1 = micromibi@meta.data$annotationV1
head(svdf)
write.csv(svdf, '../data/mibi_cell_resultann_0411.csv', quote = FALSE)
```

Now we can do cell type barplot

```{r}
full_meta = read.csv('../data/meta_mibi_dsp.csv', na.strings=c("","NA"))
colnames(full_meta)[2] = "run"
colnames(full_meta)[3] = "fov"

plot_mibi_celltype = svdf
plot_mibi_celltype = left_join(plot_mibi_celltype, full_meta, by = c('fov', 'run'))
plot_mibi_celltype$X.x <- NULL
plot_mibi_celltype$X.y <- NULL

# add more meta info
plot_mibi_celltype$tissue = str_sub(plot_mibi_celltype$run, -7,-6)
plot_mibi_celltype$treatment = str_sub(plot_mibi_celltype$tissue, -3,-2)

# also need to remove the lymph node fov
plot_mibi_celltype = plot_mibi_celltype[(plot_mibi_celltype$run != '2023-03-08T18-48-43-DSST3-Run5') & # remove a specific fov that is discarded
                                          (plot_mibi_celltype$fov != 'fov-1-scan-1'),]
```


```{r}
celltype_df <- plot_mibi_celltype %>% group_by(location,treatment, annV1) %>%
  summarise(n = n()) %>%
  mutate(freq = n / sum(n))

celltype_df$annV1 <- factor(celltype_df$annV1,
                                           levels = c("Epithelial", "Epithelial (Muc2+)" ,"Goblet", "Smooth muscle", "Smooth muscle (Ki67 high)",
                                                      "B", 'Plasma',"T", "Monocyte", "Macrophage", 'DC', 'Other Immune',
                                                      'Neutrophil'))
celltype_df = celltype_df[!is.na(celltype_df$annV1),]

pbar = ggplot(data=celltype_df, aes(x=interaction(treatment, location), y=freq, fill=annV1)) +
  geom_bar(stat="identity") + theme_classic() +
  scale_fill_manual(values = c("#8DD3C7","#BEBADA", "#FB8072", "#A6CEE3",
                                "#1F78B4", "#B2DF8A", "#33A02C",
                                "#FB9A99", "#E31A1C", "#FDBF6F",
                                "#FF7F00", "#1B9E77", "#D95F02"))

ggsave('../plots/fullBar_mibi_cell.svg', height = 3.5, width = 5) # per col 0.5, height 3.5 
ggsave('../plots/fullBar_mibi_cell.png', height = 3.5, width = 5) # per col 0.5, height 3.5 
pbar
```

Then we can calcualte the cell type log fold changes

```{r}
celltype_df_perFOV <- plot_mibi_celltype %>% group_by(run,fov,treatment,location,annV1) %>%
  summarise(n = n()) %>%
  mutate(freq = n / sum(n))
```


```{r}
# sep
celltype_df_perFOV_ctrl_small = subset(celltype_df_perFOV, (celltype_df_perFOV$treatment == 'C') &
                                         (celltype_df_perFOV$location == 'Small'))

celltype_df_perFOV_ctrl_large = subset(celltype_df_perFOV, (celltype_df_perFOV$treatment == 'C') &
                                         (celltype_df_perFOV$location == 'Large'))

celltype_df_perFOV_dss_small = subset(celltype_df_perFOV, (celltype_df_perFOV$treatment == 'T') &
                                         (celltype_df_perFOV$location == 'Small'))

celltype_df_perFOV_dss_large = subset(celltype_df_perFOV, (celltype_df_perFOV$treatment == 'T') &
                                         (celltype_df_perFOV$location == 'Large'))

# calculate per fov average of celltype composition for ctrl sample
ctrl_small_avg = celltype_df_perFOV_ctrl_small %>% group_by(annV1) %>% summarize(avg = mean(freq))
ctrl_large_avg = celltype_df_perFOV_ctrl_large %>% group_by(annV1) %>% summarize(avg = mean(freq))
dss_small_avg = celltype_df_perFOV_dss_small %>% group_by(annV1) %>% summarize(avg = mean(freq))
dss_large_avg = celltype_df_perFOV_dss_large %>% group_by(annV1) %>% summarize(avg = mean(freq))

# calculate log2 change
fc_small = dss_small_avg$avg / ctrl_small_avg$avg
fc_small = fc_small + 0.00001
fc_small = log2(fc_small)

fc_large = dss_large_avg$avg / ctrl_large_avg$avg
fc_large = fc_large + 0.00001
fc_large = log2(fc_large)

plotfc2 = data.frame(celltype = ctrl_large_avg$annV1, fc2_small = fc_small, fc2_large = fc_large)
plotfc2$celltype <- factor(plotfc2$celltype,
                        levels = c("Epithelial", "Epithelial (Muc2+)" ,"Goblet", "Smooth muscle", "Smooth muscle (Ki67 high)",
                                                      "B", 'Plasma',"T", "Monocyte", "Macrophage", 'DC', 'Other Immune',
                                                      'Neutrophil'))

#ggplot(plotfc2, aes(x=reorder(celltype, fc2_small), y=fc2_small, fill = celltype)) +
#  geom_bar(stat='identity') + theme_classic() + ggtitle("small intestine change") +
#  scale_fill_manual(values = c("#8DD3C7","#BEBADA", "#FB8072", "#A6CEE3",
#                                "#1F78B4", "#B2DF8A", "#33A02C",
#                                "#FB9A99", "#E31A1C", "#FDBF6F",
#                                "#FF7F00", "#1B9E77", "#D95F02")) #+ggtitle(celltype)
```


############ do testing and only plot out p values first


```{r}
# make new container dataframe
celltypes = unique(celltype_df_perFOV_ctrl_small$annV1)
result_ctner = data.frame(celltype = celltypes, pvalue_small = 1, pvalue_large = 1)
rownames(result_ctner) = result_ctner$celltype

location = 'Large'

ctrl_data = rbind(celltype_df_perFOV_ctrl_small, celltype_df_perFOV_ctrl_large)
dss_data = rbind(celltype_df_perFOV_dss_small, celltype_df_perFOV_dss_large)

for (celltype in celltypes){
  x = ctrl_data[(ctrl_data$location == location) & (ctrl_data$annV1 == celltype),"freq"]
  y = dss_data[(dss_data$location == location) & (dss_data$annV1 == celltype),"freq"]
  padj = t.test(y,x, paired=FALSE)$p.value #* 22 # vanilla adjusted
  result_ctner[celltype, 'pvalue_large'] = padj
}


location = 'Small'
celltypes = unique(celltype_df_perFOV_ctrl_small$annV1)

ctrl_data = rbind(celltype_df_perFOV_ctrl_small, celltype_df_perFOV_ctrl_large)
dss_data = rbind(celltype_df_perFOV_dss_small, celltype_df_perFOV_dss_large)

for (celltype in celltypes){
  x = ctrl_data[(ctrl_data$location == location) & (ctrl_data$annV1 == celltype),"freq"]
  y = dss_data[(dss_data$location == location) & (dss_data$annV1 == celltype),"freq"]
  padj = t.test(y,x, paired=FALSE)$p.value #* 22 # vanilla adjusted
  result_ctner[celltype, 'pvalue_small'] = padj
}
```


Make fold change plot with small p values

```{r}
sigctyp = result_ctner[result_ctner$pvalue_small <= 0.1, 'celltype']
temp = plotfc2[plotfc2$celltype %in% sigctyp,]

p1 = ggplot(temp, aes(x=reorder(celltype, fc2_small), y=fc2_small, fill = celltype)) +
  geom_bar(stat='identity', colour="black", linewidth = 0.2, width = 0.8) + theme_classic() + ggtitle("small intestine change") +
  ylim(c(-1, 1)) + theme(legend.position="none") + 
  scale_fill_manual(values = c("Epithelial" = "#8DD3C7",
                                "Goblet" = "#FB8072",
                                "Other Immune" = "#1B9E77",
                                "Smooth muscle (Ki67 high)" = "#1F78B4",
                                "Macrophage" = "#FDBF6F",
                                "Plasma" = "#33A02C",
                                "B" = "#B2DF8A")) 
ggsave('../plots/small_change_mibi_sig.svg', height = 3.5, width = 4.2) # per col 0.5, height 3.5 
ggsave('../plots/small_change_mibi_sig.png', height = 3.5, width = 4.2) # per col 0.5, height 3.5 
p1
```

```{r}
sigctyp = result_ctner[result_ctner$pvalue_large <= 0.1, 'celltype']
temp = plotfc2[plotfc2$celltype %in% sigctyp,]

p2 = ggplot(temp, aes(x=reorder(celltype, fc2_large), y=fc2_large, fill = celltype)) +
  geom_bar(stat='identity', colour="black", linewidth = 0.2, width = 0.8) + theme_classic() + ggtitle("large intestine change") +
  ylim(c(-1.5, 2.5)) + theme(legend.position="none") + 
  scale_fill_manual(values = c( "Macrophage" = "#FDBF6F",
                                "Plasma" = "#33A02C",
                                "DC" = "#FF7F00",
                                "Monocyte" = "#E31A1C",
                                "T" = "#FB9A99"))

ggsave('../plots/large_change_mibi_sig.svg', height = 3.5, width = 3) # per col 0.5, height 3.5 
ggsave('../plots/large_change_mibi_sig.png', height = 3.5, width = 3) # per col 0.5, height 3.5 
p2
```

generate umap plot

```{r}
umap_all = as.data.frame(micromibi@reductions$umap@cell.embeddings)
umap_all$ann = micromibi@meta.data$annotationV1
umap_all$ann <- factor(umap_all$ann,
                        levels = c("Epithelial", "Epithelial (Muc2+)" ,"Goblet", "Smooth muscle", "Smooth muscle (Ki67 high)",
                                                      "B", 'Plasma',"T", "Monocyte", "Macrophage", 'DC', 'Other Immune',
                                                      'Neutrophil'))

p = ggplot(umap_all, aes(x = UMAP_1, y = UMAP_2, color = ann)) + geom_point(size = 0.2, stroke = 0.2) +
  theme_classic() + 
  scale_color_manual(values = c("#8DD3C7","#BEBADA", "#FB8072", "#A6CEE3",
                                "#1F78B4", "#B2DF8A", "#33A02C",
                                "#FB9A99", "#E31A1C", "#FDBF6F",
                                "#FF7F00", "#1B9E77", "#D95F02")) 

ggsave('../plots/all_cell_umapV2.png', height = 6, width = 8)
p

```

